---
title: "Dealing with time series data"
author: "Noah Gallego"
date: "2024-09-18"
---

```{r, echo=FALSE}
require(mosaic)
require(ggplot2)
```


# Import the dataset into  of S&P 500  (^GSPC.csv) datset into R as sp500_data
```{r, echo=TRUE}
sp_df = read.csv("^GSPC.csv")
head(sp_df)

sp_df$Date <- as.Date(sp_df$Date, format = "%m/%d/%Y")
```

# Different time formats


We have already used some of these, and will use them again in
the format() function:

%d: Day of the month (number)

%m: Month (number)

%b: Month (abbreviated)

%B: Month (full name)

%y: year (2 digit)

%Y: year (4 digit)

%a: Weekday names (abbreviated)

%A: Weekday names (full name)

%T: Equivalent to %H:%M:%S

%u: Weekday as a decimal number (1-7, Monday is 1)

Specify date variable in the proper format

```{r, echo=TRUE}
sp_df$week_day <- format(sp_df$Date, "%A")
sp_df$Month <- format(sp_df$Date, "%B")
head(sp_df)
```

# Make a new dataframe with variables "Date", "AdjClose", "Volume"

```{r , echo=FALSE}
new_df = sp_df[c("Date", "AdjClose", "Volume")]
head(new_df)
```


# Take a subset of data from 01/01/2005 to 9/22/2023


```{r, echo=TRUE}
cat("Old DF Shape:", nrow(new_df), "\n")
new_df = new_df[new_df$Date >= "2005-01-01" & new_df$Date <= "2023-09-22", ]
cat("Number of Entries from January 2005 to September 2023:", nrow(new_df))
```

#  Calculate new variable called log_ _returns using "AdjClose" & Delete rows with "NA" for log returns

Log returns = log(P_t/P_t-1) = Log(Pt) - Log(P(t-1))
where Pt denotes today's AdjClose = Adjusted closing price
```{r, echo=TRUE}
new_df$log_returns = c(NA, diff(log(new_df$AdjClose)))
new_df$log_returns =  replace(new_df$log_returns, is.na(new_df$log_returns), mean(new_df$log_returns, na.rm = TRUE))
print(nrow(new_df))
```
# Plot "AdjClose" vs Date for the duration

```{r , echo=FALSE}
ggplot(new_df, aes(x = Date, y = AdjClose)) + 
  geom_line() +
  xlab("Date") +
  ylab("Price")

```

# Can you identify different times of economic hardships from this figure?
* 2008 Financial Crisis
* 2018 Trade War
* 2020 COVID Pandemic

# Subset data from 01/01/2018 to 09/22/2023 and draw the same diagram

```{r , echo=FALSE}
subset_df = new_df[new_df$Date >= "2018-01-01" & new_df$Date <= "2023-09-22", ]
ggplot(subset_df, aes(x = Date, y = AdjClose)) +
  geom_line() + 
  xlab("Date") +
  ylab("Price")
```

Can you identify how prices fluctuated at the beginning of the pandemic?

# Subset data from 01/01/2007 to 01/01/2023 and fit a simple linear regression model to predict AdjClose (price) using volume and draw a scatterplot with fitted line as well 


```{r , echo=FALSE}
subset_df = new_df[new_df$Date >= "2007-01-01" & new_df$Date <= "2023-01-01", ]
ggplot(subset_df, aes(x = Date, y = AdjClose)) + 
  geom_line() + 
  xlab("Date") + 
  ylab("Price")

```

# View Linear Regression Descriptive Statistics
```{r}
line = lm(AdjClose~Volume, data = subset_df)
summary(line)
```
# View Linear Regression On Plot
```{r}
# Subset data from 01/01/2007 to 01/01/2023 and fit a simple linear regression model to predict AdjClose (price) using volume and draw a scatterplot with fitted line as well 

ggplot(subset_df, aes(x = Volume, y = AdjClose)) +
  geom_point(color = "blue") + 
  geom_smooth(method = 'lm', se = FALSE, color = "red") +  
  xlab("Volume") +
  ylab("Adjusted Close Price") +
  ggtitle("Linear Regression of Adjusted Close Price on Volume (2007-2023)")
```