---
title: 'Survey Analysis: Math 3210 Project 1'
author: "Noah G. Gallego"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: true
    toc: true
    toc_depth: 2
    number_sections: true
    highlight: tango
    keep_tex: true
    latex_engine: xelatex
  word_document:
    toc: true
    toc_depth: '2'
---

# Introduction
In this project, we will analyze survey data collected from classmates to gain insights into their financial habits and situations, including budgeting methods, spending patterns, saving behaviors, and concerns about debt. We’ll conduct data cleaning and exploratory data analysis (EDA) to identify trends and patterns within financial practices. Using descriptive statistics, visualizations, and inferential analysis, we aim to highlight relationships between financial behaviors and various factors that may influence financial decision-making. This analysis will provide a deeper understanding of the financial habits and challenges among our peers.

## Load Libraries:
```{r}
library(dplyr)
library(ggplot2)
library(tm)
library(wordcloud)
library(RColorBrewer)
library(reshape2)
library(cluster)
library(ggcorrplot)
library(tidyr)
library(tidyverse)
library(stats)
library(caret)
```

## Load Data & Basic Cleaning
```{r}
# Load the DataFrame
df = read.csv("ProData.csv", header = TRUE)

# Remove Unnecessary Rows
df = df[-c(1, 2), ]

# Select Needed Cols
cols = c("RecordedDate", "LocationLatitude", "LocationLongitude", "Question.One.", "Question.Two._1", "Question.3.", "Question.4_1", "Question.5", "Question.6", "Question.7", "Question.8", "Question.9", "Question.10", "Question.11")
df = df[, cols]
df = df %>%
  rename(
    Date_Recorded = RecordedDate,
    Latitude = LocationLatitude,
    Longitude = LocationLongitude,
    Budget_Management_Method = Question.One.,
    Financial_Confidence = Question.Two._1,
    Student_Loans = Question.3.,
    Student_Debt_Concern = Question.4_1,
    Future_Financial_Decision_Impact = Question.5,
    Biggest_Expense = Question.6,
    Employment_Status = Question.7,
    Saving_For_Future_Expenses = Question.8,
    Saving_For_Retirement = Question.9,
    Financial_Independence_Confidence = Question.10,
    Job_Confidence_Post_Graduation = Question.11
  )

# Display First Few Rows
df
```

## Further Cleaning
```{r}
# Convert Cols to Ints where Needed
df = df %>%
  mutate(across(c(Latitude, Longitude, Financial_Confidence, Student_Debt_Concern), as.numeric))

# Convert Response to DT
df$Date_Recorded = as.Date(df$Date_Recorded)

# Convert Binary Responses to 1/0 For Analysis
df = df %>%
  mutate(
    Student_Loans = ifelse(trimws(tolower(Student_Loans)) == "yes", 1, 0),
    Financial_Independence_Confidence = ifelse(trimws(tolower(Financial_Independence_Confidence)) == "confident", 1, 0),
    Job_Confidence_Post_Graduation = ifelse(trimws(tolower(Job_Confidence_Post_Graduation)) == "confident", 1, 0),
    Saving_For_Future_Expenses = ifelse(trimws(tolower(Saving_For_Future_Expenses)) == "yes", 1, 0),
    Saving_For_Retirement = ifelse(trimws(tolower(Saving_For_Retirement)) == "yes", 1, 0)
  )

# Function to calculate the mode
get_mode = function(x) {
  unique_x = na.omit(x)  # Remove NA values
  unique_x[which.max(tabulate(match(x, unique_x)))]
}

# Replace NAs with mode for each column
df = df %>%
  mutate(across(everything(), ~ ifelse(is.na(.), get_mode(.), .)))

# Re-Display Dataframe
head(df)
```

## Exploratory Data Analysis (EDA)

### Descriptive Statistics
```{r}
# Summary Statistics
summary(df)

# Mean, Median, Mode, Standard Deviation for Numerical Columns
mean_values = sapply(df %>% select_if(is.numeric), mean, na.rm = TRUE)
median_values = sapply(df %>% select_if(is.numeric), median, na.rm = TRUE)
mode_values = sapply(df %>% select_if(is.numeric), get_mode)
std_dev_values = sapply(df %>% select_if(is.numeric), sd, na.rm = TRUE)

list(mean = mean_values, median = median_values, mode = mode_values, sd = std_dev_values)
```

### Missing Values Analysis
```{r}
# Check for Missing Values
missing_values = colSums(is.na(df))
print(missing_values)
```

### Visualizations

#### Financial Confidence Distribution
```{r}
ggplot(df, aes(x = Financial_Confidence)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Financial Confidence Distribution (1 = Low, 5 = High)", x = "Financial Confidence", y = "Frequency")
```

#### Employment Status Distribution
```{r}
# Bar Plot for Employment Status
ggplot(df, aes(x = as.factor(Employment_Status))) +
  geom_bar(fill = "orange", color = "black") +
  labs(title = "Employment Status Distribution", x = "Employment Status (0 = Unemployed, 1 = Employed)", y = "Count")
```

#### Financial Confidence vs. Student Debt Concern
```{r}
# Scatter Plot of Financial Confidence vs. Student Debt Concern
ggplot(df, aes(x = Financial_Confidence, y = Student_Debt_Concern)) +
  geom_point(color = "darkgreen") +
  labs(title = "Scatter Plot of Financial Confidence vs Student Debt Concern",
       x = "Financial Confidence", y = "Student Debt Concern")
```

### Additional Visualizations

#### Budget Management Method Distribution
```{r}
# Bar Plot for Budget Management Methods
ggplot(df, aes(x = Budget_Management_Method)) +
  geom_bar(fill = "purple", color = "black") +
  labs(title = "Budget Management Method Distribution", x = "Budget Management Method", y = "Count")
```

#### Saving for Future Expenses vs. Employment Status
```{r}
# Bar Plot for Saving for Future Expenses vs. Employment Status
ggplot(df, aes(x = as.factor(Employment_Status), fill = as.factor(Saving_For_Future_Expenses))) +
  geom_bar(position = "dodge", color = "black") +
  labs(title = "Saving for Future Expenses by Employment Status", x = "Employment Status (0 = Unemployed, 1 = Employed)", y = "Count", fill = "Saving for Future Expenses (0 = No, 1 = Yes)")
```

#### Financial Independence Confidence by Employment Status
```{r}
# Bar Plot for Financial Independence Confidence by Employment Status
ggplot(df, aes(x = as.factor(Employment_Status), fill = as.factor(Financial_Independence_Confidence))) +
  geom_bar(position = "dodge", color = "black") +
  labs(title = "Financial Independence Confidence by Employment Status", x = "Employment Status (0 = Unemployed, 1 = Employed)", y = "Count", fill = "Financial Independence Confidence (0 = Not Confident, 1 = Confident)")
```

### Word Cloud Analysis

#### Word Cloud for Future Financial Decision Impact
```{r}
# Create a Word Cloud for Future Financial Decision Impact
text_data = Corpus(VectorSource(df$Future_Financial_Decision_Impact))
text_data = text_data %>%
  tm_map(content_transformer(tolower)) %>%
  tm_map(removePunctuation) %>%
  tm_map(removeNumbers) %>%
  tm_map(removeWords, stopwords("english"))

wordcloud(text_data, max.words = 100, random.order = FALSE, colors = brewer.pal(8, "Dark2"))
```

## Inferential Analysis

### Regression Analysis

#### Predicting Financial Confidence
```{r}
# Linear Regression to Predict Financial Confidence
lm_model = lm(Financial_Confidence ~ Employment_Status + Student_Loans + Saving_For_Future_Expenses + Saving_For_Retirement, data = df)
summary(lm_model)
```

### Cross-Tabulations

#### Cross-Tabulation: Budgeting Method by Employment Status
```{r}
# Cross-tabulation of Budgeting Method by Employment Status
budget_vs_employment = table(df$Budget_Management_Method, df$Employment_Status)
print(budget_vs_employment)
```

#### Cross-Tabulation: Financial Confidence by Saving Habits
```{r}
# Cross-tabulation of Financial Confidence by Saving for Future Expenses
confidence_vs_saving = table(df$Financial_Confidence, df$Saving_For_Future_Expenses)
print(confidence_vs_saving)
```

### Clustering Analysis

#### K-Means Clustering of Financial Behaviors
```{r}
# K-Means Clustering
df_numeric = df %>% select_if(is.numeric)
kmeans_result = kmeans(df_numeric, centers = 3)
df$Cluster = as.factor(kmeans_result$cluster)

# Plot Clustering Results with Circles
ggplot(df, aes(x = Financial_Confidence, y = Student_Debt_Concern, color = Cluster)) +
  geom_point() +
  stat_ellipse(aes(group = Cluster), type = "norm", linetype = 2) +
  labs(title = "K-Means Clustering of Financial Behaviors", x = "Financial Confidence", y = "Student Debt Concern", color = "Cluster")
```

### Percentage Analysis of Survey Responses
```{r}
# Percentage of Responses for Key Questions

# Percentage of Students with Student Loans
total_responses = nrow(df)
students_with_loans = sum(df$Student_Loans == 1)
percentage_loans = (students_with_loans / total_responses) * 100
cat("Percentage of students with student loans:", percentage_loans, "%\n")

# Percentage of Students Saving for Future Expenses
saving_for_expenses = sum(df$Saving_For_Future_Expenses == 1)
percentage_saving_expenses = (saving_for_expenses / total_responses) * 100
cat("Percentage of students saving for future expenses:", percentage_saving_expenses, "%\n")

# Percentage of Students Saving for Retirement
saving_for_retirement = sum(df$Saving_For_Retirement == 1)
percentage_saving_retirement = (saving_for_retirement / total_responses) * 100
cat("Percentage of students saving for retirement:", percentage_saving_retirement, "%\n")

# Percentage of Employed Students
employed_students = sum(df$Employment_Status == 1)
percentage_employed = (employed_students / total_responses) * 100
cat("Percentage of students employed while attending school:", percentage_employed, "%\n")

# Percentage of Students Confident in Financial Independence
confident_financial_independence = sum(df$Financial_Independence_Confidence == 1)
percentage_confident_independence = (confident_financial_independence / total_responses) * 100
cat("Percentage of students confident in achieving financial independence:", percentage_confident_independence, "%\n")

# Percentage of Students Confident in Job Post Graduation
confident_job_post_graduation = sum(df$Job_Confidence_Post_Graduation == 1)
percentage_confident_job = (confident_job_post_graduation / total_responses) * 100
cat("Percentage of students confident in finding a job post-graduation:", percentage_confident_job, "%\n")
```

### Additional EDA

#### Heatmap of Correlations
```{r}
# Heatmap of Correlations to visualize relationships between variables
ggcorrplot(correlation_matrix, lab = TRUE, title = "Correlation Heatmap", 
           lab_size = 3,         
           tl.cex = 8,        
           tl.srt = 45,
           hc.order = TRUE)
```

#### Boxplots for Spending Habits
```{r}
# Boxplot for Biggest Expense Categories
ggplot(df, aes(x = Biggest_Expense, y = Financial_Confidence)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Boxplot of Financial Confidence by Biggest Expense", x = "Biggest Expense", y = "Financial Confidence")
```

### Geographic Analysis Using Mapview
#```{r}
#library(mapview)
# Plot the locations using mapview
#mapview(df, xcol = "Longitude", ycol = "Latitude", crs = 4269, grid = FALSE)
#```
