---
title: "Lab assignment 13 | The penguins data frame "
author: "Noah Gallego"
date: "2023-10-25"
output: word_document
---

# Read the data

```{r, echo=FALSE}
library(palmerpenguins)
library(ggthemes)
library(ggplot2)
require(mosaic)
penguins = penguins
```

# Remove all rows with NA for sex

```{r, echo=TRUE}
penguins = penguins[!is.na(penguins$sex), ]
head(penguins)
```

# Count missing values NA in each column

```{r, echo=TRUE}
penguins %>% summarise_all(~ sum(is.na(n)))
```

# Draw a Scatterplot displaying the relationship between flipper lengths and body masses of these penguins, taking into consideration the species of the penguin.

```{r, echo=TRUE}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) + 
  geom_point() +
  labs(title = "Penguin Flipper Dimensions by Species", x = "Flipper Length (mm)", y = "Body Mass (g)") +
  theme_minimal()
```

# Develop a classification rule to classify each penguin to its species

1.  Remove variables island, sex and year and call it peng.clust

```{r, echo=TRUE}
peng.clust = penguins %>%
  select(-island, -sex, -year)

head(peng.clust)
```

2.  Perform k means clustering using 3 clusters, PCA and display species vs clusters at the end

training_data = randomly select 300 rows as training data test_data = rest would be testing data

```{r, echo=TRUE}
# Perform K-Means Clustering
## Scale the Data
peng.clust = peng.clust %>%
  select(where(is.numeric))

## Divide Train/Test Data
train_indices = sample(nrow(peng.clust), 300, replace = FALSE)
training_data = peng.clust[train_indices, ]
test_data = peng.clust[-train_indices, ]

## Perform K=Means Clustering
kmeans_out = kmeans(training_data, centers = 3)

## Add Findings to DataFrame
training_data$cluster = as.factor(kmeans_out$cluster)

# Perform Principle Component Analysis (PCA)
pca_out = prcomp(training_data[, -ncol(training_data)], center = TRUE, scale. = TRUE)
training_data$PC1 = pca_out$x[, 1]
training_data$PC2 = pca_out$x[, 2]

# Plot Species vs Clusters using the PCA components
ggplot(training_data, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.6) +
  labs(title = "K-Means Clustering of Penguins - PCA Components",
       x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()
```

# Compare above diagram with the first scatterplot you generated, do you see any similarities?

-   Gentoo & Chinstrap, appear to be very similar and are harder to separate. Perhaps, the clusters could be improved from Feature Engineering.

# Make a table of cluster vs species for the test data

```{r, echo=TRUE}
find_nearest_cluster <- function(data, centers) {
  apply(data, 1, function(row) {
    distances <- apply(centers, 1, function(center) sum((row - center)^2))
    return(which.min(distances))
  })
}

# Predict clusters for the test data
test_clusters = find_nearest_cluster(test_data, kmeans_out$centers)

# Create a table of cluster vs. species
test_species = penguins$species[-train_indices]
table(Cluster = test_clusters, Species = test_species)
```

# Summarize results in the test data using a table

```{r, echo=TRUE}
test_results = data.frame(
  Species = penguins$species[-train_indices],
  Cluster = as.factor(test_clusters)
)

summary_table = table(test_results$Cluster, test_results$Species)
print(summary_table)
```

# Take the penguins datset after remove NA rows for sex and perform a PCA

## How many PCA's should we conside according to the screeplot

```{r, echo=TRUE}
# Perform PCA on the numeric columns of penguins after removing NAs for sex
peng.pca = prcomp(peng.clust, center = TRUE, scale. = TRUE)

# Scree plot to visualize explained variance
explained_variance = peng.pca$sdev^2 / sum(peng.pca$sdev^2)
plot(explained_variance, type = "b", 
     xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained",
     main = "Scree Plot for PCA")

```

# Draw a stacked bar plots to visualize the relationship between two categorical variablesisland and species

```{r, echo=TRUE}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "stack") +
  labs(title = "Distribution of Penguin Species by Island",
       x = "Island",
       y = "Count",
       fill = "Species") +
  theme_minimal()
```

# Draw a frequncy table between two categorical variablesisland and species

```{r, echo=TRUE}
freq_table = table(penguins$island, penguins$species)
print(freq_table)
```

# Following diagram shows the distribution of different species in the three islands, modify the code shows the distribution of different species and islands among males and females

```{r, echo=TRUE}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species)) +
  facet_grid(sex ~ island) +
  labs(title = "Distribution of Species by Island and Sex",
       x = "Flipper Length (mm)", y = "Body Mass (g)") +
  theme_minimal()
```
