---
title: "Lab 4: RMarkdown and RStudio Overview"
author: "Noah Gallego"
date: "09 September 2024"
output: 
  html_document: 
    toc: false
    df_print: paged
  word_document: 
    toc: false
  pdf_document: 
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import Packages
```{r}
library(plotrix)
library(readxl)
library(VennDiagram)
library(dplyr)
library(tidyr)

```

## Question 1: Load the Excel file
```{r}
# Read-In Excel File
df = read_xlsx("survey.xlsx")
```

## Question 2: Convert all text in the data frame to lowercase
```{r}
# Convert All Text in df to lowercase
for (i in colnames(df)) {
  df[[i]] = tolower(df[[i]])
}
```

## Question 3: Split the expectation column into two new columns
```{r}
# Split expectation column into two separate columns
df = separate(df, expectation, into = c("expectation_1", "expectation_2"), sep = ",")

# Remove blank spaces
df$expectation_2 = trimws(df$expectation_2)
```

## Question 4: Replace "na" in expectation_1 with NA values
```{r}
# Replace "na" with NA
df = mutate(df, mutate(across(everything(), ~ na_if(., "na"))))
```

## Question 5: Create a frequency table for motivation
```{r}
# Create a Frequency Table for motivation
plot_freq = function(plotted_df, title, x, y = "Frequency") {
  barplot(plotted_df,
          main = title,
          xlab = x,
          ylab = y,
          col = "lightblue"
  )
}

motivation_table = table(df$motivation)
plot_freq(motivation_table,
          "Motivation for Taking the Class",
          "Motivation"
          )
```

## Question 6: Create a frequency table for R proficiency levels
```{r}
# Create a Frequency Table for Knowledge Levels
k_table = table(df$r_level)
plot_freq(k_table,
          "Self-Assessed Knowledge Prior to Class",
          "Knowledge Levels"
)
```

## Question 7: Generate frequency tables for both expectations
```{r}
# Create Frequency Tables for Expectations
df$expectation_1 = gsub("_", " ", df$expectation_1, fixed=TRUE)
e_1 = table(df$expectation_1)
e_2 = table(df$expectation_2)
data = rbind(e_1, e_2)

barplot(data,
        beside = TRUE,
        main = "Expectations for Taking the Class",
        xlab = "Expectation",
        ylab = "Frequency",
        col = c("lightblue", "lightgreen"),
        names.arg = c("Data Analyst", "R Coder"),
        legend.text = c("Data Analyst", "R Coder")
)
```

## Question 8: Venn diagram of overlap between expectations
```{r}
# Create a Venn diagram showcasing the overlap between students who chose "data analyst" for expectation_1 and "r coder" for expectation_2
data_analyst_students = !is.na(df$expectation_1) & df$expectation_1 == "data analyst"
r_coder_students = !is.na(df$expectation_2) & df$expectation_2 == "r_coder"

venn.plot = venn.diagram(
  x = list(
    "Data Analyst" = which(data_analyst_students),
    "R Coder" = which(r_coder_students)
  ),
  category.names = c("Data Analyst", "R Coder"),
  fill = c("blue", "green"),
  alpha = 0.5,
  filename = NULL
)
df
grid.draw(venn.plot)

```

## Question 9: Proportion of each R proficiency level as a pie chart
```{r}
# Proportion of R proficiency levels
k_table_prop = prop.table(k_table)
pie3D(k_table_prop,
      labels=unique(df$r_level),
      explode=0.5,
      main="Proportion of Knowledge Levels",
      labelcex = 1.5
)
```

## Question 10: Cross-tabulation between R level and motivation
```{r}
# Cross-Tabulation between R Level and Motivation
cross_tab <- table(df$r_level, df$motivation)
print(cross_tab)
```

## Question 11: Stacked bar plot for cross-tabulation of R level and motivation
```{r}
# Generate a Stacked Bar Chart to visualize cross-tabulation
barplot(cross_tab, 
        main = "Stacked Bar Chart of R Level and Motivation", 
        col = c("red", "green", "blue"), 
        legend = rownames(cross_tab), 
        xlab = "Motivation", 
        ylab = "Frequency"
)
```

## Question 12: Frequency distribution of motivation in a bar plot
```{r}
# Create a Frequency Table for Motivation
motivation_table = table(df$motivation)
plot_freq(motivation_table,
          "Motivation for Taking the Class",
          "Motivation"
)
```

## Question 13: Proportion of each motivation value in a pie chart
```{r}
# Proportion of Motivation Values
motivation_prop = prop.table(motivation_table)
pie3D(motivation_prop,
      labels=unique(df$motivation),
      explode=0.5,
      main="Proportion of Motivation for Taking Class",
      labelcex = 1.5
)
```
