---
title: "Reshaping Your Data with the tidyr | Lab assignment 8"
author: "Noah Gallego"
date: "2024-09-25"
output: word_document
---

# Loading the installed package to your R session

This is the process of loading and attaching packages or libraries to your R session.

## If you're installing a package for the first time install.packages("dplyr")

```{r, echo=FALSE}
require(tidyr)
require(dplyr)
```

# Question 1

The “produceWide.txt” dataset contains the information
about yields of produce in all seasons. Reshape the dataset so the
new data frame will have three columns: ID, season, and yield,
where the variable season has four levels (spring, summer, fall and
winter).

Read the data set as produceWide

```{r, echo=TRUE}
df_wide = read.table("produceWide.txt", header = TRUE)
head(df_wide)
```

```{r, echo=TRUE}
df_long = df_wide %>% 
  pivot_longer(cols = c(Spring, Summer, Fall, Winter),
               names_to = "season",
               values_to = "yields")
head(df_long)
```


# Question 2

The “produceLong.txt” dataset contains the information
about yields of produce in all seasons. Read it as produceLong and Reshape the produceLong data frame to wide format.


```{r, echo=TRUE}
produce_df_long = read.table("produceLong.txt", header = TRUE)
head(produce_df_long)
```


```{r, echo=TRUE}
produce_df_wide = produce_df_long %>%
  pivot_wider(names_from = season,
              values_from = yield)
head(produce_df_wide)
```




# Question 3

Read the S&p 500 index data from "^GSPC.csv". Separate the "Date" column into three columns as "month","day","year"

```{r, echo=TRUE}
sp_df = read.csv("^GSPC.csv")
sp_df = sp_df %>%
  separate(col = Date,
           into = c("month", "day", "year"),
           sep = "/")
head(sp_df)
```


# Question 4

Read the "Diamonds.txt" file as Diamonds_data and create a new column price_weight by uniting the “WEIGHT” and
“PRICE” columns by "/" .

```{r, echo=TRUE}
diamonds_data = read.table("Diamonds.txt", header = TRUE)
diamonds_data  = diamonds_data %>%
  unite(col = price_weight, WEIGHT, PRICE, sep = "/")

head(diamonds_data)
```

# Question 5

Read the "student_performance_missing.xlsx" file as student_data. You need "readxl" package for this. Install it first

```{r, echo=TRUE}
library(readxl)
student_data = read_excel("student_performance_missing.xlsx")
head(student_data)
```


Now replace missing values in Exam_score with mean of the Exam_score values

```{r, echo=TRUE}
mean_exam_score = mean(student_data$Exam_Score, na.rm = TRUE)

student_data <- student_data %>%
  mutate(Exam_Score = ifelse(is.na(Exam_Score), mean_exam_score, Exam_Score))

head(student_data)
```

