---
title: "Graphical and Numerical summaries"
author: "Noah Gallego"
date: "2024-09-11"
output: pdf_document
---
# Read Data
data = read.csv("BBgradrates.csv")
head(data)

install.packages("tidyverse")
library(readxl)
data_excel = read_xlsx("BBgradrates.xlsx")

head(data_excel)
# Dotplot example using R


## Installing necessary R packages


## Loading the installed package to your R session

This is the process of loading and attaching packages or libraries to your R session.


```{r, echo=FALSE}
require(mosaic)
```

# Sactterplot 

## Import the data set to R

Import the dataset into R
```{r, echo=TRUE}
data("trees")
```

## Checking the dataset overview
```{r, echo=TRUE}
help(trees) #Gives a detailed description of the data set "Trees". Look at the window in your right hand bottom corner
glimpse(trees)
```

## Plotting the scatter plot

```{r , echo=FALSE}
help("xyplot")
xyplot( Volume ~ Height, data = trees, main = "Scatterplot of Height vs Volume", xlab = "Height", ylab = "Volume" )

```


# Dot plot

## Import the data set to R

Import the dataset into R
```{r, echo=TRUE}
gradrate = read.csv("BBgradrates.csv")
```

## Checking the dataset overview
```{r, echo=TRUE}
glimpse(gradrate)
```

## Plotting the dot plot

```{r , echo=FALSE}
gf_dotplot( ~ Graduation_rates , data = gradrate)
```


## Comparative Dotplots

R syntax:  gf_dotplot( ~ x | gfactor , data).


```{r, echo=TRUE}
SeatBelt_data = read.csv("SeatBeltStudy.csv")
glimpse(SeatBelt_data)
gf_dotplot( ~ Perc_wearing | Law , data=SeatBelt_data)
```

# Calculating mean using r

Import the dataset into R
```{r, echo=TRUE}
movies_revenue = read.csv("movies_revenue.csv")
```



## mean

```{r, echo=TRUE}
mean( ~ revenue, data= movies_revenue )
```


# Calculating median using r

```{r, echo=TRUE}
median( ~ revenue, data= movies_revenue )
```


# Calculating quantiles

## Quantile 1
```{r, echo=TRUE}
 quantile(movies_revenue$revenue, 0.25)
```


## Quantile 2
```{r, echo=TRUE}
 quantile(movies_revenue$revenue, 0.50)
```

## Histogram 
```{r, echo=TRUE}
options(scipen=999)
histogram( ~ revenue, data= movies_revenue )
```


# Histogram for The HELP study

Import the dataset to R

```{r, echo=TRUE}
HstudyData = read.csv("HELPstudy.csv")
```

Glimpse of the data set

```{r, echo=TRUE}

glimpse( HstudyData )


```
Histogram for age


```{r, echo=TRUE}
histogram( ~ age , data= HstudyData )
```

Comparative histogram to examine the distribution of age by the type of housing (homeless or housed). The housing type observations are in the housing variable.


```{r, echo=TRUE}
histogram( ~ age | housing , data= HstudyData )
```

# Measures of dispersion and Boxplot

## Import cholestrol datset

```{r, echo=TRUE}
choldata  = read.csv("chol.csv")
```

## Glimpse of dataset

```{r, echo=TRUE}
glimpse(choldata)
```


## Variance of cholestrol

variance: var( ~ chol , data = choldata)

```{r, echo=TRUE}
var( ~ chol , data = choldata)
```

### Variance by group
```{r, echo=TRUE}
var( ~ chol|group , data = choldata)
```

## Standard deviation
```{r, echo=TRUE}
sd ( ~ chol , data = choldata)
```

### Standard deviation by group
```{r, echo=TRUE}
sd ( ~ chol | group , data = choldata)
```


## Median cholestrol level
```{r, echo=TRUE}
median( ~ chol , data = choldata)
```


## Median cholestrol level by group
```{r, echo=TRUE}
median( ~ chol | group , data = choldata)
```

###  IQR of colestrol level
```{r, echo=TRUE}
iqr( ~ chol , data = choldata)
```

## Five number summary
```{r, echo=TRUE}
fivenum( ~ chol , data = choldata)
```

### Five number summary by group
```{r, echo=TRUE}
fivenum( ~ chol , data = choldata)
```

## Box plot
```{r, echo=TRUE}
gf_boxplot( ~ chol , data = choldata)
```


###  Box plot by group
```{r, echo=TRUE}
gf_boxplot( ~ chol | group , data = choldata)
```

# Robust statistics

Following are some MATH 2200 marks entered to canvas by your instructor
```{r, echo=TRUE}
robust_data = c(88,87,86,91,79,89,7,85,900,90,92,96,85,6,850)
```

Calculate mean, median, range, IQR, standard deviation, variance

```{r, echo=TRUE}
mean(robust_data)
median(robust_data)
range(robust_data)
iqr(robust_data)
sd(robust_data)
var(robust_data)
```

## Draw a boxplot for the above data

```{r, echo=TRUE}
boxplot(robust_data)
```

## Remove the outliers

Lower outliers
```{r, echo=TRUE}
Q1 =  quantile(robust_data,0.25)
IQR = iqr(robust_data)
lower_out = Q1 - 1.5*IQR 
lower_out
```

Upper outliers
```{r, echo=TRUE}
Q3 =  quantile(robust_data,0.75)
IQR = iqr(robust_data)
upper_out = Q3 + 1.5*IQR 
upper_out
```

## Now remove the outliers

```{r, echo=TRUE}
robust_data_withoutoutliers = c(88,87,86,91,79,89,85,90,92,96,85)
```

Calculate mean, median, range, IQR, standard deviation, variance

```{r, echo=TRUE}
mean(robust_data_withoutoutliers)
median(robust_data_withoutoutliers)
range(robust_data_withoutoutliers)
iqr(robust_data_withoutoutliers)
sd(robust_data_withoutoutliers)
var(robust_data_withoutoutliers)
```


## Draw a boxplot for the above data

```{r, echo=TRUE}
boxplot(robust_data_withoutoutliers)
```
