---
title: "Lab Seven"
author: "Noah Gallego"
date: "2024-09-29"
output: word_document
---

# Advanced Dplyr Assignment with Multiple Excel Files

### Problem 1: Data Manipulation (PeopleData)
```{r setup, echo = FALSE}
# Load Packages
library(dplyr)
library(tidyr)
library(readxl)
```

```{r echo = FALSE}
# Load Data
people_data = read_excel("../Data/PeopleData.xlsx")
transaction_data = read_excel("../Data/TransactionsData.xlsx")
employee_data = read_excel("../Data/EmployeeData.xlsx")

# Create new df
df1 = people_data %>% 
  select(age, sex, race, avg_drinks_perday)

# Create new Age Group Column Based on Age
df1$age_group = cut(df1$age,
  breaks = c(-Inf, 30, 50, Inf),
  labels = c("Young", "Adult", "Senior"))

# Create new Drinker Status Column Based on Avg Drinks / Day
df1$drinker_status = cut(df1$avg_drinks_perday,
                         breaks = c(-Inf, 0, 1, 3, Inf),
                         labels = c("Non-Drinker", "Light Drinker", "Moderate Drinker", "Heavy Drinker")
                         )

# Filter seniors who are heavy drinkers & calculate the average age
df1 %>% filter(age_group == "Senior", `avg_drinks_perday` > 3) %>%
  group_by(sex) %>%
  
  summarize(
    avg_age = mean(age, na.rm = TRUE),
    total_count = n()
  )
```

### Problem 2: TransactionsData and EmployeeData Analysis
```{r}
head(transaction_data)

# Group Transaction Data By Category & Sum
transaction_data %>% 
  group_by(category) %>%
  summarize(
    total_amount = sum(amount, na.rm = TRUE),
    average_amount = mean(amount, na.rm = TRUE)
  )
```

```{r}
# Join the Transaction Data w/ Employee Data 
#head(employee_data)
#merged_data = transaction_data %>%
# left_join(employee_data, by = "employee_id")

# There is no common column??
```
### Problem 3: Advanced Grouping and Ranking
```{r}
people_data$age_group = df1$age_group

# Group By Age and Race
grouped_data = people_data %>%
  group_by(age_group, race) %>%
  summarise(
    avg_age = mean(age, na.rm = TRUE),
    total_people = n()
  )

# Rank Individuals by avg_drinks_perday within each age_group
ranked_data = people_data %>%
  group_by(age_group) %>%
  mutate(rank = dense_rank(desc(avg_drinks_perday))) %>%
  filter(rank <= 3)

ranked_data = ranked_data[order(ranked_data$rank), ]

# Display the top three individuals in each group
ranked_data
```

### Problem 4: Joining Data & Complex Summaries
```{r}
## Perform an inner-join between PeopleData and TransactionsData using a common column
#joined_data = people_data %>%
# inner_join(transaction_data, by = "customer_id")
# No common column???? If the data was correct, it would look like:

## Calculate total transaction amount for Heavy Drinkers
# total_transaction_amount = joined_data %>%
#  summarise(
#    total_amount = sum(case_when(
#      avg_drinks_perday > 3 ~ amount,
#      TRUE ~ 0
#    ), na.rm = TRUE)
#  )


# avg_transaction_by_race = joined_data %>%
#  filter(avg_drinks_perday > 3) %>%
#  group_by(race) %>%
#  summarise(avg_amount = mean(amount, na.rm = TRUE))
```

### Problem 5: Pollution Characteristics
```{r}
pollution_df = read.table("../Data/pollution1-1.txt", header = TRUE)

str(pollution_df)
```
### Problem 6: Pollution Data Selection & Cleaning
```{r}
# Clean DataFrame
pollution_df$date = as.Date(pollution_df$date, format = "%Y-%m-%d")
pollution_df = pollution_df %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

# Use select() to keep every variable that ends with a two
pollution_df_four = pollution_df %>%
  select(ends_with("2"))

# Display the first few rows
head(pollution_df_four)
```

### Problem 7: The Filter Function
```{r}
# Filter by a day, i.e the world championship day
champion_df = pollution_df %>% 
  filter(date == "2005-10-23")

# Select Temp & Dew Point from date
temp_info = champion_df %>%
  select(tmpd, dptp)
temp_info

# Filter PM2.5 levels greater than 30 and temperature that is greater than 80
filtered_pm_temp = pollution_df %>%
  filter(pm25tmean2 > 30, tmpd > 80)
filtered_pm_temp

# Count the number of days that satisfy the condition
num_days = nrow(filtered_pm_temp)
num_days

# See what months are in the filtered data
library(lubridate)
filtered_pm_temp %>%
  mutate(month = month(date)) %>%
  distinct(month)
```

### Problem 8: Arrange Dates in Descending Order
```{r}
# Arrange Dates in Descending Order
df_arranged = pollution_df %>%
  arrange(desc(date))

head(df_arranged)
```

### Problem 9: Create a Year Variable and Display It
```{r}
# Create year Variable
df_with_year = pollution_df %>%
  mutate(year = year(date))

head(df_with_year)
```

### Problem 10: Group Data By Year & Compute Median of O3 Levels
```{r}
df_median_o3 = df_with_year %>%
  group_by(year) %>%
  summarise(median_o3 = median(o3tmean2, na.rm = TRUE))

head(df_median_o3)
```